Slightly Positive Correlation Between Racialization, Gendering and Police Strip Searches in the Toronto Municipal, 2020-2021

### 2024-01-26 outputs/paper/paper.qmd
update dicussion and graphs

---
title: "Slightly Positive Correlation Between Racialization, Gendering and Police Strip Searches in the Toronto Municipal, 2020-2021"
title-block-banner: true
abstract: "Police strip search has been a topic of great contention during the COVID-19 pandemic. The question commonly raised was how certain demographics are more likely to undergo strip searches than others, and why has this skewed power dynamics been maintained for so long. Data then was gathered from Open Data Toronto to assess the correlation between strip searches and the racial and gender identities of arrested 'suspects' in the Toronto Municipal from 2020 to 2021. In short, this paper finds that there is a slight positive correlation between police strip searches and the gendered and racialized Toronto demographic such as females or people of African descent. This analysis would enable us to rethink the modern form of policing, how it transpires in times of crisis and critically affects people of marginalized identities." 
thanks: "Code and data are available at: https://github.com/ponolite/Police_Arrests_and_Strip_Searches/tree/main"
author: "Quang Mai"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: references.bib
---

@@

Recently, the Ford government has been at the forefront of public push-back in Toronto, due to their revisions of strip search laws under the Ministry of Correctional Services Act, R.S.O. 1990, c. M.22. Addressed by the Canadian Civil Liberties Association, the revisions "fall short...to ensure that prisoners are treated with basic respect" [@ontarioStripLaw]. By definition, strip search is a police enforcement, which entails removing some or all articles of clothing to visually inspect a person's private body parts and see if they harbor any items like weapons or drugs [@torontoPolice]. Surprisingly, while contended as a dehumanizing policing act, little has been done to assess strip search as a form of police power abuse in Toronto. This is especially relevant considering the city's racialized and gendered demographics, how strip searches most likely affect these social groups under the subtext of COVID-19 from 2020 to 2021 [@stripSearchOpEd].

This paper thus inspects gender and racial identities of all Toronto police arrests from 2020 to 2021, and assesses if there is a correlation between marginalized identities and police strip searching. Foremost, this paper aims to be a form of data activism, contributing to the dialogue around police power abuse that enforces strip-search to undermine the the vulnerable populations in Toronto. To analyze, I first map out the distribution of Toronto police arrests by race and gender and reasons for strip searching from 2020 to 2021 using summary tables, stacked histograms and bubble charts. Particularly, for each category of race and gender, the distribution bars or points would be sectioned into various categories like 'including strip searches' or 'not including' to visualize the statistics of strip searching against intersecting social identities. It's worth noting that this dataset may already contain much racial and gender bias, because it was retrieved from the very source that this paper is questioning: the Toronto Police Services [@stripSearchOpEd]. Then, the findings presented might not fully encapsulate the actual landscape of strip searching in Toronto from 2020 to 2021. Nonetheless, it's still worth looking at due the precedents of police power abuse [@covidStripSearch].

@@

### Racialized and Gendered Police Strip Searches Statistics

To encapsulate, throughout 2020 and 2021, there are a total of 7801 recorded strip searches out of 65276 incidents of police arrests, which accounts for a whooping 12% of all arrests (see [@tbl-items-strip-search]). Excluding those whose racial identities are unknown and non-existent, by percentage, the three most affected racial groups are, in descending order, of Indigenous, Black and White ethnic roots, standing at 16%, 14% and 13% each (see [@tbl-race-strip-search]). Gender-wise, men are more likely to be strip searched compared to women, at 12% and 10% respectively (see [@tbl-gender-strip-search]). Though, observably, the difference is quite marginal. To note, people of unidentified gender identity aren't strip searched at all according to the dataset (denoted as `U` in the tables and graphs).

@@

Particularly, accounting for the discrepancies between counts of strip search grouped by different racial and gender identities are challenging. This is due to the incongruous population spread of each racial and gender group within Toronto. For instance, a major 44.3% of Toronto population is white people [@ethnicCensus], thus it's more probable that the people arrested and strip searched by the police force are largely `white` in the dataset (see [@@tbl-race-strip-search]) As such, to render the data statistically proportionate to one another, `mutate()`was largely used to create new variables that are mostly percentage values. Admittingly so, this enables each variable to be on common grounds when compared with one another.

In addition, a new column called "Total Reasons" totals all reasons used to account for each police strip search is added with `mutate()`, where all instances of `1` and `0` are changed into numeric with `as.numeric()` and then summed together.
to ensure that all data are comparative to one another

```{r fig.width = 8}
#| echo: false
#| message: false
#| label: fig-race_strip_search
#| fig-cap: Racial Identities and Police Strip Search Per Arrest in Toronto, 2020-2021

@@

@@

```{r fig.height=3, fig.width=8}
#| echo: false
#| message: false
#| label: fig-gender_strip_search
#| fig-cap: Gender Identities and Police Strip Search Per Arrest in Toronto, 2020-2021

cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(count = n) |>
  group_by(gender) |>
  mutate(percentage = round(count/sum(count)*100)) |>
  ggplot(mapping = aes(x = strip_search, fill = strip_search), data=cleaned_race_gender_data) +
  geom_bar() + 
  theme_minimal() +
  labs(y = "Count", x = "Strip Search", fill = "Strip Search") +
  facet_wrap(~gender) + scale_fill_viridis_d(option="cividis")

As for reasons of strip searching provided by the Toronto Police Services and items found along these strip searches, there is a substantial 10% of strip searches being conducted without any reason (see [@tbl-reasons-strip-search]). Most commonly, strip searches are carried out under at least 1 or 2 reasons, both at 29% (see [@tbl-reasons-strip-search]). Overall, most of strip search enforcement doesn't garner any items in question from the people who were suspected by the police. Specifically, a whooping 63% of strip searches don't have any items found (see [@tbl-items-strip-search]).

@@

cleaned_search_reasons_data |>
  group_by(total_reasons, items_found) |>
  count() |>
  rename(count = n) |>
  group_by(total_reasons) |>
  mutate(percentage = round(count/sum(count)*100),
         total_searches = sum(count),
         percentage_total = round(total_searches/7801*100)) |>
  kable(
    col.names = c("Total Reasons", "Items Found", "Count", "Percentage by Reasons", "Total Searches", "Percentage by Searches"),
    booktabs = TRUE,
    align = c("c", "c", "c", "c", "c", "c")
  ) |> 
  kable_styling(latex_options="scale_down")
```

```{r fig.height=4, fig.width=8}
#| echo: false
#| message: false
#| label: fig-reasons-race
#| fig-cap: Relationship Between Reasons for Strip Search and Race in Toronto, 2020-2021
#| fig-subcap: ["Race","Gender"]
#| layout-ncol: 2

#create a scatterplot to show each observation in the dataset, comparing the total reasons to racial identitiy found in each strip search
#colour the points based on the perceived race of the captive
#code for geom_point() referenced from https://github.com/TDonofrio62/Shooting-Occurrences-in-Toronto/blob/main/Shooting_Occurrences/outputs/paper/Shooting_Occurrences.Rmd
cleaned_search_reasons_data |>
  group_by(total_reasons, race) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = total_reasons, y = count, colour = race)) +
  geom_line() +
  geom_point(aes(size = count)) +
  labs(x = "Total Reasons for Strip Searching",
       y = "Count",
       color = "Race"
       ) +
  theme_minimal() +
  theme(legend.key.size = unit(0.5, 'cm')) +
  scale_fill_viridis_d(option="cividis")

cleaned_search_reasons_data |>
  group_by(total_reasons, gender) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = total_reasons, y = count, colour = gender)) +
  geom_line() +
  geom_point(aes(size = count)) +
  labs(y = "Count", x = "Total Reasons for Strip Searching", color = "Gender") +
  theme(legend.key.size = unit(0.5, 'cm')) +
  theme_minimal() +
  scale_fill_viridis_d(option="cividis")

```{r fig.height=3, fig.width=8}
#| echo: false
#| message: false
#| results: hide
#| label: fig-reasons-gender
#| fig-cap: Relationship Between Reasons for Strip Search and Gender in Toronto, 2020-2021

cleaned_search_reasons_data |>
  group_by(total_reasons, gender) |>
  count() |>
  rename(count = n) |>
  ggplot(aes(x = total_reasons, y = count, colour = gender)) +
  geom_line() +
  geom_point(aes(size = count)) +
  labs(y = "Count", x = "Total Reasons for Strip Searching", color = "Gender") +
  theme(legend.key.size = unit(0.5, 'cm')) +
  theme_minimal() +
  scale_fill_viridis_d(option="cividis")

## Discussion

binaristic view of gender

## Conclusion

>>>>

---
title: "Slightly Positive Correlation Between Racialization, Gendering and Police Strip Searches in the Toronto Municipal, 2020-2021"
title-block-banner: true
abstract: "Police strip search has been a topic of great contention during the COVID-19 pandemic. The question commonly raised was how certain demographics are more likely to undergo strip searches than others, and why has this skewed power dynamics been maintained for so long. Data then was gathered from Open Data Toronto to assess the correlation between strip searches and the racial and gender identities of arrested 'suspects' in the Toronto Municipal from 2020 to 2021. In short, this paper finds that police strip searches are slightly racialized and gendered, especially towards people of African descent and females within Toronto. This analysis would enable us to rethink the modern form of policing, how it transpires in times of crisis and critically affects people of marginalized identities." 
thanks: "Code and data are available at: https://github.com/ponolite/Police_Arrests_and_Strip_Searches/tree/main"
author: "Quang Mai"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: references.bib
---

@@

Recently, the Ford government has been at the forefront of public push-back in Toronto, due to their revisions of strip search laws under the Ministry of Correctional Services Act, R.S.O. 1990, c. M.22. Addressed by the Canadian Civil Liberties Association, the revisions "fall short...to ensure that prisoners are treated with basic respect" [@ontarioStripLaw]. By definition, strip search is a police enforcement, which entails removing some or all articles of clothing to visually inspect a person's private body parts and see if they harbor any items like weapons or drugs [@torontoPolice]. Surprisingly, while contended as a dehumanizing policing act, little has been done to assess strip search as a form of police power abuse in Toronto. This is especially relevant considering the city's racialized and gendered demographics [@ethnicCensus; @genderCensus], how strip searches most likely affect these social groups under the subtext of COVID-19 from 2020 to 2021 [@stripSearchOpEd].

This paper thus inspects gender and racial identities of all Toronto police arrests from 2020 to 2021, and assesses if there is a correlation between marginalized identities and police strip searching. Foremost, this paper aims to be a form of data activism, contributing to the dialogue around police power abuse that enforces strip-search to undermine the the vulnerable populations in Toronto. To analyze, I first map out the distribution of Toronto police arrests by race and gender and reasons for strip searching from 2020 to 2021 using summary tables, stacked bar charts and bubble charts. Particularly, for each category of race and gender, the distribution bars or points would be sectioned into various categories like 'including strip searches' or 'not including' to visualize the statistics of strip searching against intersecting social identities. It's worth noting that this dataset may already contain much racial and gender bias, because it was retrieved from the very source that this paper is questioning: the Toronto Police Services [@stripSearchOpEd]. Then, the findings presented might not fully encapsulate the actual landscape of strip searching in Toronto from 2020 to 2021. Nonetheless, it's still worth looking at due the precedents of police power abuse [@covidStripSearch].

@@

### Racialized Police Strip Searches Statistics

To encapsulate, throughout 2020 and 2021, there are a total of 7801 recorded strip searches out of 65276 incidents of police arrests, which accounts for a whooping 12% of all arrests (see [@tbl-items-strip-search]). Excluding those whose racial identities are unknown and non-existent, by percentage, the three most affected racial groups are, in descending order, of Indigenous, black and white background, standing at 16%, 14% and 13% each (see [@tbl-race-strip-search]).

@@

Particularly, accounting for the discrepancies between counts of strip search grouped by different racial and gender identities are challenging. This is due to the incongruous population spread of each racial and gender group within Toronto. For instance, 44.3% of Toronto population is made up of white people [@ethnicCensus], thus it's more probable that the people arrested and strip searched by the police force are ethnically "White" in the dataset (see [@tbl-race-strip-search]). As such, to render the data statistically proportionate to one another, `mutate()`was used to create new variables that are mostly percentage values.

Then, @fig-race-strip-search illustrates the percentage of strip searches among different racial groups in Toronto from 2020-2021. Observably, while being the the major population group in Toronto, people of white racial roots aren't the most strip searched. Rather, it's the racialized population of Toronto like black and Indigenous people that are being strip searched by police. This hints at the possibility of strip search being more racialized than it seems, targeting minority groups in Toronto.

```{r fig.height = 4, fig.width = 8}
#| echo: false
#| message: false
#| label: fig-race-strip-search
#| fig-cap: Racial Identities and Police Strip Search Per Arrest in Toronto, 2020-2021

@@

### Gendered Police Strip Searches Statistics

Gender-wise, men are more likely to be strip searched compared to women, at 12% and 10% respectively (see [@tbl-gender-strip-search]). Though, observably, the difference is quite marginal. To note, people of unidentified gender identity aren't strip searched at all according to the dataset (denoted as `U` in the tables and graphs).

@fig-gender-strip-search illustrates the percentage of gender groups being strip searched in Toronto. It's observable that the number of strip searches between men and women are comparable, especially considering the 48-52% population split between the two categories of gender in Toronto, according to the 2021 Census [@genderCensus].

Nonetheless, the dataset only contains a binaristic view of gender, without regarding those who identify as non-binary or others, and essentially lumping all other gender identities as "Unidentified". This proves inconsistent against the constant public backlash agasint police for their power abuse of people belonging to the LGBTQI+ community in Toronto [@policeGender]. This point will be further discussed below.

```{r fig.height=3, fig.width=5}
#| echo: false
#| message: false
#| label: fig-gender-strip-search
#| fig-cap: Gender Identities and Police Strip Search Per Arrest in Toronto, 2020-2021

cleaned_race_gender_data |>
  group_by(gender, strip_search) |>
  count() |>
  rename(count = n) |>
  group_by(gender) |>
  mutate(percentage = round(count/sum(count)*100)) |>
  ggplot(aes(x = gender, y = percentage, fill = strip_search)) + 
  geom_bar(stat="identity", position = "dodge") +
  theme_minimal() +
  theme(legend.key.size = unit(0.5, 'cm')) +
  labs(x = "Race",
       y = "Percentage",
       fill = "Strip Search") + scale_fill_viridis_d(option="cividis")

In general, there is a substantial 10% of strip searches being conducted without any reasons (see [@tbl-reasons-strip-search]). Commonly, strip searches are carried out by at least 1 or 2 reasons provided by the police, both at 29% (see [@tbl-reasons-strip-search]). As for items found during searches, at a whooping 63%, most strip search enforcement doesn't garner any suspected items (see [@tbl-items-strip-search]). This critically denotes how many strip searches aren't motivated by civic considerations but are moreso carried out by intuition, or personal biases that can extend to racial profiling. It signifies an alarming potential for misconducted strip searches that are racially or sexually motivated. To visualize this data, @fig-reasons-race and @fig-reasons-gender both map, respectively, the racial and gender distribution of strip searches by the total number of reasons each strip search has.

Regarding @fig-reasons-race, due to the sheer number of police searches that people who are racially black and white have compared to other groups, this paper subsections the dataset on racialized police searches into two graphs. Both of these graphs are devoid of data from people are racially unknown or don't identify with any races--"None" or "Unknown or Legacy"--to ensure visibility. In specifics, graph a) of @fig-reasons-race contains only the distribution from people who racially identify as black and white, graph b) of @fig-reasons-race contains the distribution of all other minority groups--except for black people--against white people. This ensures data clarity between groups with higher counts of strip searches and groups with lower counts of strip searches. In addition, to ensure that data is proportionate to one another, instead of using `count`, the percentages for each racial group categorized by the different sums of reasons within their total number of strip searches are calculated.

Based on part a) of @fig-reasons-race, the number of black people who are strip searched surpass white people on all counts of total number of reasons, except for instances where the police only has one reason to conduct strip search. This signifies an alarming rate of racialized strip searches against black people despite their small population make-up in Toronto (@ethnicCensus). For part b) of @fig-reasons-race, the percentages between other minority groups fluctuate. But intriguingly, compared to white people, all of other minority groups are more likely to be strip searched with only  Groups that are most commonly strip searched without any provided reasons are "Middle-Eastern", "East/Southeast Asian" and "Indigenous", at 11%, 11% and 10% each.

Population-wise,

```{r fig.height=4, fig.width=8}
#| echo: false
#| warning: false
#| message: false
#| label: fig-reasons-race
#| fig-cap: Relationship Between Reasons for Strip Search and Race in Toronto, 2020-2021
#| n-cols: 2
#| fig-subcap: ["Black and White","Other Minority Groups", "new"]
#| layout-ncol: 3

#create a scatterplot to show each observation in the dataset, comparing the total reasons to racial identitiy found in each strip search
#colour the points based on the perceived race of the captive
#code for geom_point() referenced from https://github.com/TDonofrio62/Shooting-Occurrences-in-Toronto/blob/main/Shooting_Occurrences/outputs/paper/Shooting_Occurrences.Rmd

cleaned_search_reasons_data |>
  filter(race == "Black" | race == "White") |>
  group_by(total_reasons, race) |>
  count() |>
  rename(count = n) |>
   group_by(race) |>
  mutate(percentage = round(count/sum(count)*100)) |>
  ggplot(aes(x = total_reasons, y = percentage, color = race)) +
  geom_point(aes(size = percentage)) +
  geom_smooth() +
  labs(x = "Total Reasons for Strip Searching",
       y = "Percentage",
       color = "Race"
       ) +
  theme_minimal() +
  theme(legend.key.size = unit(0.5, 'cm'))

cleaned_search_reasons_data |>
  filter(race != "None" & race != "Unknown or Legacy" & race != "Black") |>
  group_by(total_reasons, race) |>
  count() |>
  rename(count = n) |>
  group_by(race) |>
  mutate(percentage = round(count/sum(count)*100)) |>
  ggplot(aes(x = total_reasons, y = percentage, color = race)) +
  geom_point(aes(size = percentage)) +
  geom_smooth() +
  labs(x = "Total Reasons for Strip Searching",
       y = "Percentage",
       color = "Race"
       ) +
  theme_minimal() +
  theme(legend.key.size = unit(0.5, 'cm')) 
```

```{r fig.height=3 , fig.width=8}
#| echo: false
#| message: false
#| warning: false
#| label: fig-reasons-gender
#| fig-cap: Relationship Between Reasons for Strip Search and Gender in Toronto, 2020-2021

cleaned_search_reasons_data |>
  group_by(total_reasons, gender) |>
  count() |>
  rename(count = n) |>
   group_by(gender) |>
  mutate(percentage = round(count/sum(count)*100)) |>
  ggplot(aes(x = total_reasons, y = percentage, colour = gender)) +
  geom_point(aes(size = percentage)) +
  geom_smooth() +
  labs(y = "Count", x = "Total Reasons for Strip Searching", color = "Gender") +
  theme(legend.key.size = unit(0.5, 'cm')) +
  theme_minimal() +
  scale_fill_viridis_d(option="cividis")

## Discussion and Conclusion

'None' and unknown legacy, and unidentified
difficult work of categorizing
binaristic view of gender

@@

Overall, the data reveals relationships between demographic factors (age, gender, and area or neighborhood) and number of MHA apprehensions reports. By conducting histogram graphs, We found that the number of MHA reports increases each year until 2022: in comparison, the total crime reports in Toronto decreases until 2020 (@fig-years). This trend of data could be a result of Covid-19 Pandemic due to the fact that many people encounter mental health issues while staying mostly at homes. The difference of the patterns in MHA and total crime reports also hints a potential indirect correlation with each other. Demographically, males in age cohort from 25 to 34 are more likely to be reported as MHA apprehensions(`r male_per`% males, `r age_per`% with age cohort from 25 to 34) as shown in @fig-groups. The difference between gender and age groups are not necessarily significant though are most likely to be since the large sample size. Closer studies such as model and power analysis should be conducted for better statistical understanding of the data sets.

Additionally, an possible indirect relation between the number of MHA apprehension and corresponding area division is detected: increase in number of total crime in one division does not correlates to higher number of MHA apprehensions with the same degree within areas (@fig-merged). The percentages of average MHA reports out of average police reports of the years are not consistent across the divisions. In other words, possible bias, mediating, or moderating factors could take place in the correlation equation such as @eq-relation. Although the paper does not create and testify the validity of the equations and models, it provides a potential future direction for the research field.

While these findings point towards a slight positive correlation between police strip search being racially inclined and gendered, particularly towards the black and female population of Toronto, it's worth noting how the data has decontextualized the reasons the police account for each strip search. Specifically, all reasons-- including "Injury", "Weapons", "Escape" and "Possess Evidence"--are quantified as a single unit for each arrest, without further details on the severity scale of the situation. In short, all of these reasons are weighed equally. This can very much seclude the fact that some arrests--even  are more necessary to be strip searched than others, thus skewing the nature of this paper's analysis. For instance, reffering back to @tbl-reasons-strip-search, the more number of reasons to account for a strip search doesn't necessarily mean that a person is bound to be strip searched. The number of reasons the dataset attributes to each strip search--as provided by Toronto Police Service [@torontoPolice]--can be purely projected based on the police in power's racial preconceptions or gender biases. This means that a person with only one reason to be strip searched may be more "necessary" to be strip searched than others. Nonetheless, this paper believes that the categorization of the reasons this way by the dataset somewhat confirms th

That said, there are further pain-points regarding the nature of the dataset, making the analysis suggestive in nature. As mentioned, noting the combative nature between the Toronto police force and people from the LGBTQI+ community [@policeGender], this paper recognizes that simplifying the gender spectrum to a mere binary of males and females can disregard the struggles of non-binary people (see [@tbl-gender-strip-search]). This problem of 'categorization' is also problematic for the ".

Additional socio-economic measures can further ascertain the validity of this dataset

by summing these reasons together. Specifically, the severity of the reasons aren't factored within this paper analysis, which can hinder its conclusion. For instance, a person involved in the strip search might have caused grave injuries or possessed dangerous

Although these findings point to inequity in child care access based on average income per ward, many other factors must also be taken into consideration. The child care data provided by the Child Service's division only contains information about licensed child care centres [@caredata]. The Government of Ontario specifies that unlicensed home child care providers can operate with five or fewer children [@ontario]. This may account for many of the children who are unable to obtain a space in a licensed child care facility as may family members and private sitters, but this data is unavailable through Open Data Toronto. The average cost of child care at each facility could help to provide more insight into equitable access but this information is also unavailable in this data set.

A variable which may skew the data or produce outliers is the fact that each of Toronto's wards contain multiple neighborhoods with various social and economic considerations [@wards]. Outliers such as Toronto Centre and Scarborough Rouge-Park for instance may be influenced by variation in income by neighborhood or higher child care supply based on commuting population and business density. Although average household income by ward was used in @fig-childrenspacesplot and @fig-childrenspacesscatterplot, an analysis using the median household income of each ward showed a similar trend (see Appendix: @fig-spacesmedianbywards & @fig-spacesmedianincome).

## Appendix {-}

### Strip Search and Items Found Correlation

@tbl-reasons-strip-search connotes the summary statistics how many reasons the police force attributes to each person prior to conducting strip searches and the number of items found during each search. This table leverages `mutate` to create three new columns on the data's percentages and counts, enabling the data to be proprotionate to one another. "Total Reasons" conveys the total number of reasons a strip search has, while "Items Found Percentage" conveys the percentage of items found for each total of reasons. "Total Searches by Total Reasons" conveys the total number of searches for each total of reasons, while "Searches Percentage" conveys the percentage of each total of reasons out of all strip searches.

cleaned_search_reasons_data |>
  group_by(total_reasons, items_found) |>
  count() |>
  rename(count = n) |>
  group_by(total_reasons) |>
  mutate(percentage = round(count/sum(count)*100),
         total_searches = sum(count),
         percentage_total = round(total_searches/7801*100)) |>
  kable(
    col.names = c("Total Reasons", "Items Found", "Count", "Items Found Percentage", "Searches by Total Reasons", "Searches Percentage"),
    booktabs = TRUE,
    align = c("c", "c", "c", "c", "c", "c")
  ) |> 
  kable_styling(latex_options="scale_down")
```

@@
